\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
OS comments

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function. As a consequence, the focus has been on reporting hazard ratios instead of survival curves. Indeed, reporting patient risk requires a separate estimation of the baseline hazard. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression. Their approach naturally leads to estimates of the survival function that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

The semiparametric Cox model has become the default approach to survival analysis 
even though Cox himself later suggested he would prefer to model the hazard function 
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
how he would model a set of censored survival data, to which he responded: "I 
think I would normally want to tackle problems parametrically ... and if you want 
to do things like predict the outcome for a particular patient, it's much more 
convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
relevant quantity in a clinical setting is often the 5- or 10-year risk of 
experiencing a certain event given the patient's particular circumstances. 
Unfortunately, the most reported metric from a Cox model is the (potentially 
time-dependent) hazard ratio (HR), which ignores the duration of follow-up and 
is subject to selection bias [@hernan2010hazards]. The covariate-adjusted 
survival curve overcomes these limitations, and it is arguably a more important 
summary measure to report than the HR. While stepwise survival curves can be 
computed with the Cox model, they require a second step to separately estimate 
the baseline hazard [@breslow1972discussion].

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth survival curves much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical, Section 4.8]. 
On the other hand, a coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates ("person-moments") in 
continuous time from all observed follow-up experience constituting the 
study base. By doing so, they obtained a likelihood expression for the hazard 
function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements and extends the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate-adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
includes extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve, and their standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more end users with the hope that they will favor reporting absolute risks and survival curves
over hazard ratios.

In what follows, we first recall some theoretical details on case-base sampling 
and its use for estimating parametric hazard functions. We then give a short 
review of existing R packages that implement comparable features as \pkg{casebase}. 
Next, we provide some details about the implementation of case-base sampling in 
our package, and we give a brief survey of its main functions. This is followed 
by three case studies that illustrate the flexibility and capabilities of 
\pkg{casebase}. We show how the same framework can be used for non-linear 
functions of time and non-proportional hazards, competing risks, and variable 
selection via penalized regression. Finally, we end the article with a 
discussion of the results and of future directions.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*. These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile, an exposure status, and an outcome status (i.e. whether the event happened) attached to it. By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.

To complement the case series, we sample person-moments for the base series. To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series. We note that the same individual can contribute multiple person-moments to the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual likelihood contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating function for $\theta$ as follows:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, their sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base (e.g. the sum of all individual follow-up times). More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. Non-uniform sampling mechanisms can increase the efficiency of the case-base estimators.

@@

## Competing-risk analysis

Case-base sampling can also be used in the context of competing-risk analysis. Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form

$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{survival} provides functionality for competing-risk analysis and multistate modelling. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIs.

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

$$ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
From this, we can derive the event-specific subdensities:

@@

By integrating these subdensities, we obtain the event-specific CI functions:

@@

In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model non-constant and non-proportional hazards through a flexible specification of time. Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

@@

$$
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
$$
where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the 
variable $t$ (which is usually time, but can be any other continuous variable), 
$x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for 
the unexposed group, $\mathbf{Z}$ are other covariates in the model which are 
equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, 
and $\hat{\beta}$ are the estimated regression coefficients. As indicated by the 
formula above, it is most instructive to plot the hazard ratio as a function of 
a variable $t$ only if there is an interaction between $t$ and $X$. Otherwise, 
the resulting plot will simply be a horizontal line across time.

@@

In this case study, we show how case-base sampling can be used in the context of a competing-risk analysis. For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      2001 McGill College Avenue
      Montreal, QC, Canada H3A 1G1
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      2001 McGill College Avenue
      Montreal, QC, Canada H3A 1G1
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      2001 McGill College Avenue
      Montreal, QC, Canada H3A 1G1
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function. As a consequence, the focus has been on reporting hazard ratios instead of survival curves. Indeed, reporting patient risk or cumulative incidence requires a separate estimation of the baseline hazard. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression. Their approach naturally leads to estimates of the survival or risk function that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can also be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

The semiparametric Cox model has become the default approach to survival analysis 
even though Cox himself later suggested he would prefer to model the hazard function 
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
how he would model a set of censored survival data, to which he responded: "I 
think I would normally want to tackle problems parametrically ... and if you want 
to do things like predict the outcome for a particular patient, it's much more 
convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
relevant quantity in a clinical setting is often the 5- or 10-year risk of 
experiencing a certain event given the patient's particular circumstances. 
However, the most reported metric from a Cox model is the (potentially 
time-dependent) hazard ratio (HR). The covariate-conditional
survival curve is arguably a more important 
summary measure to report than the HR. While stepwise survival curves can be 
computed with the Cox model, they require a second step to separately estimate 
the baseline hazard [@breslow1972discussion].

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth survival curves more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time; Carstensen [-@copenhagen2012needs] provides an exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical, Section 4.8], and large long-format datasets.
On the other hand, a coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates ("person-moments") in 
continuous time from all observed follow-up experience constituting the 
study base. By doing so, they obtained a likelihood expression for the hazard 
function that is equivalent to that of logistic regression with an offset term. 
More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including functions of time as covariates 
(e.g. using splines or generalized additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements and extends the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate-conditional survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
includes extensions to other models such as penalized 
regression for variable selection and competing risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve, and their standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to end users who favor reporting absolute risks and survival curves
over hazard ratios.

In what follows, we first recap some theoretical details on case-base sampling 
and its use for estimating parametric hazard functions. We then give a short 
review of existing R packages that implement comparable features as \pkg{casebase}. 
Next, we provide some details about the implementation of case-base sampling in 
our package, and we give a brief survey of its main functions. This is followed 
by three case studies that illustrate the flexibility and capabilities of 
\pkg{casebase}. We show how the same framework can be used for non-linear 
functions of time and non-proportional hazards, competing risks, and variable 
selection via penalized regression. Finally, we end the article with a 
discussion of the results and of future directions.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*. These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile and an outcome status (i.e. whether the event happened) attached to it. By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of independent censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.

To complement the case series, we sample person-moments for the base series. To do so, we specify the base series sampling mechanism as non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $\,\textrm dR_i(t) = 1$ constitute the base series. We note that the same individual can contribute multiple person-moments to the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. As mentioned above, the processes $R_i(t)$ are specified by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging using standard software.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, an estimating function for $\theta$ can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, their sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled base series *person-moments*, and $B$ is the total population-time for the study base (e.g. the sum of all individual follow-up times). More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. Non-uniform sampling mechanisms can increase the efficiency of the resulting maximum partial likelihood estimators.

@@

## Competing risk analysis

Case-base sampling can also be used in the context of competing risk analysis. Assuming there are $J$ competing events, we can show that each sampled person-moment's contribution to the partial likelihood is of the form

$$\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{survival} provides functionality for competing risk analysis and multistate modelling. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIs.

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

$$ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
From this, we can derive the cause-specific subdensities:

@@

By integrating these subdensities, we obtain the cause-specific CI functions:

@@

In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model non-constant and non-proportional hazards through a flexible specification of time. Then we perform a competing risk analysis and compare our results with the Cox model and the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

@@

$$
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
$$
where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the 
variable $t$ (which is usually time, but can be any other continuous variable), 
$x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for 
the unexposed group, $\mathbf{Z}$ are other covariates in the model which are 
equal to $\mathbf{z}$ in both the exposed and unexposed group, 
and $\hat{\beta}$ are the estimated regression coefficients. As indicated by the 
formula above, it is most instructive to plot the hazard ratio as a function of 
a variable $t$ only if there is an interaction between $t$ and $X$. Otherwise, 
the resulting plot will simply be a horizontal line across time.

@@

In this case study, we show how case-base sampling can be used in the context of a competing risk analysis. For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
